---
title: "Reproducible Research Course Project 1"
author: "Yisu Zhou"
date: '2018-04-24'
output:
    html_document:
    keep_md: true
---

## Set up knitr
Note: only included for peer evaluation purpose

```{r setup, include = TRUE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
knitr::opts_chunk$set(fig.path = 'figure/', fig.width = 8, fig.height = 6)
```

## Loading and preprocessing the data
```{r}
# should start at ~/dropbox/course/jhu-reproducible_research
library(dplyr)

fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

# datadir <- here("data")
# destfile <- here("data", "project1.zip")
if(!file.exists("./data/project1.zip")){
    res <- tryCatch(download.file(fileURL,
                                  destfile="./data/project1.zip",
                                  method="curl"),
                    error=function(e) 1)
 }

unzip("./data/project1.zip", exdir = "./data", overwrite = TRUE)

pamd <- read.table("./data/activity.csv", header = TRUE, sep = ",", na.strings="NA", stringsAsFactors=F)

```

## What is mean total number of steps taken per day?
1. Calculate the total number of steps taken per day

ANSWER:
```{r, results='asis'}
library(xtable)
tstep <- pamd %>% group_by(date) %>% summarise(tstep = sum(steps, na.rm = TRUE))
print(xtable(tstep), type = "html", na.print = "NA")

pamd$date <- as.Date(pamd$date)
```

2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day

```{r}
library(ggplot2)
qplot(tstep$tstep, geom = "histogram", xlab = "Total steps taken each day")
```

3. Calculate and report the mean and median of the total number of steps taken per day

ANSWER: Mean & median: 

```{r}
mean(tstep$tstep, na.rm = TRUE)
median(tstep$tstep, na.rm = TRUE)
```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. 𝚝𝚢𝚙𝚎 = "𝚕") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
avgstep <- pamd %>% group_by(interval) %>% summarize (avgs = mean(steps, na.rm = TRUE))

with(avgstep, plot(interval, avgs, type = "l"))
```

2. Which 5minute interval, on average across all the days in the dataset, contains the maximum number of steps?
ANSWER:

```{r}
print(avgstep[which.max(avgstep$avgs),])
```

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as 𝙽𝙰). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with 𝙽𝙰s)
	+ ANSWER:

```{r}
sum(is.na(pamd$steps))
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

ANSWER: impute missing value with the average steps of that interval period across dates

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
pamd2 <- pamd %>% group_by(interval) %>% mutate_at(vars(steps), funs(ifelse(is.na(.), mean(steps, na.rm = TRUE), .)))
```	

4. Make a histogram of the total number of steps taken each day and calculate and report the **mean** and **median** total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

ANSWER: Yes, judging by the two histograms, their distributions are different. Notably, after imputation, both the mean and median values are higher compared to un-imputed data. We can also tell that if we do not explicitly tell r to account for missing data in the original histogram (i.e. by `na.rm = TRUE`), the NA values are treated as "0", which inflates the left tail.  

```{r,results='asis'}
tstep2 <- pamd2 %>% group_by(date) %>% summarise(tstep = sum(steps, na.rm = TRUE), 
	mean = mean(steps, na.rm = TRUE), median = median(tstep, na.rm = TRUE))
tstep2$date <- as.character(tstep2$date)
print(xtable(tstep2), type = "html")
qplot(tstep2$tstep, geom = "histogram", xlab = "Total steps taken each day")
```

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
pamd2$week <- ifelse(weekdays(pamd2$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
```

2. Make a panel plot containing a time series plot (i.e. 𝚝𝚢𝚙𝚎 = "𝚕") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r}
avgstep2 <- pamd2 %>% group_by(interval, week) %>% summarize(avgs = mean(steps, na.rm = TRUE))
qplot(interval, avgs, data=avgstep2, facets = week~., geom=c("line"),
	xlab = "5-minute interval", ylab = "average steps")

```
